\documentclass[aspectratio=169]{beamer}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{hyperref}
\setbeamertemplate{navigation symbols}{}
\title{Designing and deploying a FAIR-by-design data pipeline and platform for electron microscopy laboratories}
\subtitle{Research thesis in: Data Management}
\author{Nicola Perin \and Supervisor: Dott. Federica Bazzocchi}
\institute{UNIVERSITY OF TRIESTE\\ Department of Mathematics, Informatics and Geosciences\\ Master’s Degree in Data Science and Scientific Computing\\ A.Y. 2024–2025}
\date{2025-09-19}
\begin{document}
\begin{frame}
\titlepage
\end{frame}

\begin{frame}{Agenda}
\tableofcontents
\end{frame}

\section{Context \& Goals}
\begin{frame}{Motivation \& Problem}
\begin{itemize}
\item EM labs generate very large, heterogeneous datasets (images, diffraction, spectra).
\item Vendor-specific formats + sparse metadata hinder interoperability \& reuse.
\item Informal practices (file names, local notes) don't scale across collaborations.
\end{itemize}
\end{frame}

\begin{frame}{FAIR Principles (Goal)}
\begin{itemize}
\item Make data Findable, Accessible, Interoperable, Reusable.
\item Emphasize rich metadata, standardized formats, machine actionability.
\item Align with funder \& journal expectations; enable open science.
\end{itemize}
\end{frame}

\begin{frame}{Standards \& Formats}
\begin{itemize}
\item HDF5: performant hierarchical container for large arrays + attributes.
\item NeXus: community conventions (NXinstrument, NXsample, etc.).
\item NXem: EM application definition (images, diffraction, EDS/EELS, 4D-STEM + metadata).
\end{itemize}
\end{frame}

\begin{frame}{Institutional Context}
\begin{itemize}
\item Area Science Park → RIT institute with labs: LADE (data engineering), LAGE (genomics), LAME (electron microscopy).
\item ORFEO data center provides HPC, Ceph storage, and identity services.
\item Project targets LAME workflows; scalable across NFFA‑DI.
\end{itemize}
\end{frame}

\section{Infrastructure \& Platform}
\begin{frame}{Infrastructure: ORFEO}
\begin{itemize}
\item Ceph-based distributed storage (tiers for speed/capacity; replication/erasure coding).
\item RADOS Gateway exposes S3-compatible object storage.
\item Identity \& SSO: FreeIPA directory + Authentik (OIDC).
\end{itemize}
\end{frame}

\begin{frame}{Application Platform}
\begin{itemize}
\item Web app unifies upload + structured annotation in one flow.
\item Django (MVT) + PostgreSQL; API for scripted ingest.
\item Separation of interactive deposit (lab) from heavy processing (data center).
\end{itemize}
\end{frame}

\begin{frame}{Deployment \& Validation}
\begin{itemize}
\item VirtualOrfeo digital twin: realistic, safe testbed.
\item K3s Kubernetes, NGINX Ingress, MetalLB, cert-manager.
\item Optional Ceph object storage integration; production‑like SSO.
\end{itemize}
\end{frame}

\section{Design \& Implementation}
\begin{frame}{Domain Model \& Data Flow}
\begin{itemize}
\item Project → Proposal → Sample → Experiment → Measurement hierarchy.
\item Metadata captured early; mapped into NeXus/NXem.
\item Curated outputs stored in centralized data lake (S3/Ceph).
\end{itemize}
\end{frame}

\begin{frame}{Metadata \& NeXus Construction}
\begin{itemize}
\item Automated mapping of instrument settings (beam energy, detectors, stage coords).
\item Human‑readable README + machine‑readable NeXus containers.
\item Libraries (e.g., pynxtools-em) lower adoption barrier.
\end{itemize}
\end{frame}

\begin{frame}{Storage Gateway \& Background Tasks}
\begin{itemize}
\item Upload pipeline writes to object storage; lifecycle managed server‑side.
\item Background workers handle conversion, validation, indexing.
\item Auditability \& provenance preserved end‑to‑end.
\end{itemize}
\end{frame}

\begin{frame}{Security, Performance, Scalability}
\begin{itemize}
\item OIDC tokens with group claims → role-based access in-app.
\item Ceph scales horizontally; PostgreSQL tuned for concurrent users.
\item Asynchronous jobs decouple ingestion from heavy compute.
\end{itemize}
\end{frame}

\section{Results \& Conclusions}
\begin{frame}{Contributions}
\begin{itemize}
\item FAIR-by-design workflow from acquisition to curated NeXus/NXem.
\item Unified upload + annotation web app; API surface for automation.
\item Deployment blueprint validated in VirtualOrfeo; ready to port to ORFEO/NFFA‑DI.
\end{itemize}
\end{frame}

\begin{frame}{Conclusions \& Next Steps}
\begin{itemize}
\item Reproducible, interoperable EM data pipeline proved feasible.
\item Scale to additional instruments and labs; expand validators \& viewers.
\item Prepare for open access portals and cross‑lab discovery.
\end{itemize}
\end{frame}

\end{document}